Enforced Invariant Confluence

### 2018-02-26 doc/paper_redux/iconfluence.tex
Show labels.

\begin{claim}\clmlabel{state_irec_implies_reachable}
  $\Irec{e} \implies \reachable{e}$
\end{claim}
\begin{elidableproof}
  We perform structural induction on $e$.
  \begin{itemize}
    \item \textbf{Case $s_0$.}
      Axiomatically, $\reachable{s_0}$.

@@

\begin{claim}
  Consider a state based object $(S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$. The following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

    \item
      $I(s_0)$ and $\Irec{e_1} \land \Irec{e_2} \implies I(e_1 \join e_2)$
  \end{enumerate}
\end{claim}
\begin{elidableproof}
  First, we show that (1) implies (2). Axiomatically, $\reachable{s_0}$, so by
  (1), $I(s_0)$. Let $e_1$ and $e_2$ be arbitrary expressions such that
  $\Irec{e_1}$ and $\Irec{e_2}$. By \clmref{state_irec_implies_reachable},
  $\reachable{e_1}$ and $\reachable{e_2}$. Thus, $\reachable{e_1 \join e_2}$,
  so by (1), $I(e_1 \join e_2)$.

@@

\begin{claim}\clmlabel{op_irec_implies_reachable}
  $\Irec{e} \implies \reachable{e}$
\end{claim}
\begin{elidableproof}
  We perform a structural induction on $e$.
  \begin{itemize}
    \item \textbf{Case $s_0$.}
      Axiomatically, $\reachable{s_0}$.

@@

\begin{claim}
  Given an operation-based object $(S, s_0)$, a set of transactions $T$, and an
  invariant $I$, the following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

    \item
      $I(s_0)$ and $\Irec{e_1} \land \Irec{e_2} \land I(t(e_1)(e_1)) \implies
      I(t(e_1)(e_2))$.
  \end{enumerate}
\end{claim}
\begin{elidableproof}
  First, we show that (1) implies (2). $\reachable{s_0}$, so by (1), $I(s_0)$.
  Let $e_1$ and $e_2$ be arbitrary expressions such that $\Irec{e_1}$,
  $\Irec{e_2}$, and $I(t(e_1)(e_1))$. By \clmref{op_irec_implies_reachable},
  $\reachable{e_1}$ and $\reachable{e_2}$, so $\reachable{t(e_1)(e_2)}$. By
  (1), $I(t(e_1)(e_2))$.

>>>>

\begin{claim}\clmlabel{StateBasedIrecImpliesReachable}
  $\Irec{e} \implies \reachable{e}$
\end{claim}
\begin{elidableproof}
  We perform structural induction on $e$.
  \begin{itemize}
    \item \textbf{Case $s_0$.}
      Axiomatically, $\reachable{s_0}$.

@@

\begin{claim}\clmlabel{StateBasedTwoIconfluenceDefs}
  Consider a state based object $(S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$. The following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

    \item
      $I(s_0)$ and $\Irec{e_1} \land \Irec{e_2} \implies I(e_1 \join e_2)$
  \end{enumerate}
\end{claim}
\begin{elidableproof}
  First, we show that (1) implies (2). Axiomatically, $\reachable{s_0}$, so by
  (1), $I(s_0)$. Let $e_1$ and $e_2$ be arbitrary expressions such that
  $\Irec{e_1}$ and $\Irec{e_2}$. By \clmref{StateBasedIrecImpliesReachable},
  $\reachable{e_1}$ and $\reachable{e_2}$. Thus, $\reachable{e_1 \join e_2}$,
  so by (1), $I(e_1 \join e_2)$.

@@

\begin{claim}\clmlabel{OpBasedIrecImpliesReachable}
  $\Irec{e} \implies \reachable{e}$
\end{claim}
\begin{elidableproof}
  We perform a structural induction on $e$.
  \begin{itemize}
    \item \textbf{Case $s_0$.}
      Axiomatically, $\reachable{s_0}$.

@@

\begin{claim}\clmlabel{OpBasedTwoIconfluenceDefs}
  Given an operation-based object $(S, s_0)$, a set of transactions $T$, and an
  invariant $I$, the following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

    \item
      $I(s_0)$ and $\Irec{e_1} \land \Irec{e_2} \land I(t(e_1)(e_1)) \implies
      I(t(e_1)(e_2))$.
  \end{enumerate}
\end{claim}
\begin{elidableproof}
  First, we show that (1) implies (2). $\reachable{s_0}$, so by (1), $I(s_0)$.
  Let $e_1$ and $e_2$ be arbitrary expressions such that $\Irec{e_1}$,
  $\Irec{e_2}$, and $I(t(e_1)(e_1))$. By \clmref{OpBasedIrecImpliesReachable},
  $\reachable{e_1}$ and $\reachable{e_2}$, so $\reachable{t(e_1)(e_2)}$. By
  (1), $I(t(e_1)(e_2))$.

### 2018-02-26 doc/paper_redux/interactive_decision_procedure.tex
Show labels.

\section{Interactive \Iconfluence{} Decision Procedure}
\clmref{sufficientandnecessary} arms us with some potent understanding of
\Iconfluence! Imagine we have an \Iclosed{} decision procedure that informs us
that our state-based object from \exampleref{statebasednotnecessary} is not
\Iclosed{}. Without \clmref{sufficientandnecessary}, we'd be a bit stuck. We'd
have to seek out alternative means to determine whether or not our transactions
were \Iconfluent{}. With the intuition provided by
\clmref{sufficientandnecessary}, we can instead restrict our invariant to $I =
\setst{(x, y)}{xy \leq 0 \land x \geq 0 \land y \leq 0}$. Now, the \Iclosed{}
decision procedure informs us that we are indeed \Iclosed{}, so we know we're
\Iconfluent{}.

>>>>

\section{Interactive \Iconfluence{} Decision Procedure}
Let us revisit \exampleref{statebased_not_necessary}

% \clmref{sufficientandnecessary} arms us with some potent understanding of
% \Iconfluence! Imagine we have an \Iclosed{} decision procedure that informs us
% that our state-based object from \exampleref{statebasednotnecessary} is not
% \Iclosed{}. Without \clmref{sufficientandnecessary}, we'd be a bit stuck. We'd
% have to seek out alternative means to determine whether or not our transactions
% were \Iconfluent{}. With the intuition provided by
% \clmref{sufficientandnecessary}, we can instead restrict our invariant to $I =
% \setst{(x, y)}{xy \leq 0 \land x \geq 0 \land y \leq 0}$. Now, the \Iclosed{}
% decision procedure informs us that we are indeed \Iclosed{}, so we know we're
% \Iconfluent{}.
%

### 2018-02-26 doc/paper_redux/sufficient_conditions.tex
Show labels.

\begin{claim}\clmlabel{statebasedsufficient}
  Given a state-based object $O = (S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$, if $O$ is \Iclosed{}, then $T$ is \Iconfluent.
\end{claim}
\begin{elidableproof}
  Let $e_1, e_2$ be expressions such that $\Irec{e_1}$ and $\Irec{e_2}$. Let
  $s_1 = eval(e_1)$ and $s_2 = eval(e_2)$. Then, $I(s_1)$ and $I(s_2)$, so
  $I(s_1 \join s_2)$ which implies $I(e_1 \join e_2)$. Thus, by
  \clmref{statebased_two_iconfluence_defs}, $T$ is \Iconfluent{}.
\end{elidableproof}

\begin{claim}\clmlabel{opbasedsufficient}
  Given a operation-based object $O = (S, s_0)$, a set of transactions $T$, and
  an invariant $I$, if $O$ is \Iclosed{}, then $T$ is \Iconfluent.
\end{claim}
\begin{elidableproof}
  Let $t \in T$ and $e_1, e_2$ be expressions such that $\Irec{e_1}$,
  $\Irec{e_2}$, and $I(t(e_1)(e_1))$. Let $s_1 = eval(e_1)$ and $s_2 =
  eval(s_2)$. Then, $I(s_1)$, $I(s_2)$, and $I(t(s_1)(s_1))$, so
  $I(t(s_1)(s_2))$. This implies $I(t(e_1)(e_2))$, so by
  \clmref{opbased_two_iconfluence_defs}, $T$ is \Iconfluent.
\end{elidableproof}

As a special case, \clmref{statebasedsufficient} tells us that a set $T$ of
transactions is \Iconfluent{} if $I$ is monotonic and $O$ is a semilattice.

\begin{claim}\clmlabel{monotonicity}
  Let $O = (S, s_0, \join)$ be a state-based object where $(S, \join)$ is a
  semilattice. Let $s \leq s' \defeq s \join s' = s'$. Let $T$ be an arbitrary
  set of transactions. Let $I$ be a \defword{monotonic}: for all $s, s'\in S$,
  if $s \leq s'$ and $I(s)$, then $I(s')$. Also assume $I(s_0)$. $T$ is
  \Iconfluent{}.
\end{claim}
\begin{elidableproof}
  Let $s_1$ and $s_2$ be arbitrary states such that $I(s_1)$ and $I(s_2)$. $s_1
  \leq s_1 \join s_2$, so by monotonicity, $I(s_1 \join s_2)$. Thus, by
  \clmref{statebasedsufficient}, $T$ is \Iconfluent{}.
\end{elidableproof}

Unfortunately, the converses of \clmref{statebasedsufficient} and
\clmref{opbasedsufficient} are not true. Being \Iconfluent{} does not
necessarily imply being \Iclosed{}, as the next two examples show.

\begin{example}\examplelabel{statebasednotnecessary}
  Consider the state-based object $O = (S, s_0, \join)$ where $(S, \join)$ is
  $(\ints, \max) \times (\ints, \max)$ and $s_0 = (0, 0)$. Let $T$ consist of
  two transactions $t_{x+1} \defeq \lambda (x, y).\ (x + 1, y)$ and $t_{y-1} =
  \lambda (x, y).\ (x, y - 1)$. Let $I = \setst{(x, y)}{xy \leq 0}$.
  %
  $\setst{(x, y)}{\reachable{(x, y)}} = \nats \times -\nats \subseteq I$, so
  $T$ is \Iconfluent. However, let $s_1 = (-1, 1)$ and $s_2 = (1, -1)$.
  $I(s_1)$ and $I(s_2)$. However, $s_1 \join s_2 = (1, 1)$, and $\lnot I((1,
  1))$. Thus, $O$ is not \Iclosed{}.
\end{example}

@@

Let's take a closer look at \exampleref{statebasednotnecessary} (illustrated in
\figref{statebasednotnecessary}) to understand why being \Iconfluent{} doesn't
imply being \Iclosed{}. Our counterexample considers points $s_1 = (-1, 1)$ and
$s_2 = (1, -1)$. $s_2$ is reachable, but $s_1$ is not. This is not a
coincidence! If we consider an \Iconfluent{} set of transactions, then any two
reachable points satisfy the invariant, and the join of the two points is again
reachable, so it must satisfy the invariant as well. The only way to join two
invariant satisfying points and end up not satisfying the invariant is if one
or both points are not reachable. Thus, if we know that $I$ is a subset of
reachable points, then being \Iconfluent{} and being \Iclosed{} are equivalent.

\begin{claim}\clmlabel{sufficientandnecessary}
  If every invariant satisfying state is reachable, then being \Iconfluent{}
  and being \Iclosed{} are equivalent. Being \Iclosed{} is always sufficient,
  but if we can establish that every invariant satisfying point is reachable,
  then being \Iclosed{} is a both necessary and sufficient for being
  \Iconfluent{}.
  \[
    I \subseteq \setst{e}{\reachable{e}} \implies
    (\text{\Iclosed{}} \iff \text{\Iconfluent})
  \]
\end{claim}
\begin{elidableproof}
  Consider a state-based object $O = (S, s_0, \join)$, a set of transactions
  $T$, and an invariant $I$ where $I \subseteq \setst{e}{\reachable{e}}$.
  %
  Being \Iclosed{} always implies being \Iconfluent{} whether or not $I
  \subseteq \setst{e}{\reachable{e}}$. Thus, we only have to prove that if $T$
  is \Iconfluent{}, then $O$ is \Iclosed. If $T$ is \Iconfluent{}, then
  $I(s_0)$. Next, let $s_1$ and $s_2$ be arbitrary states such that $I(s_1)$
  and $I(s_2)$. Both states satisfy the invariant, so both are reachable. That
  is, there exist expressions $e_1$ and $e_2$ where $eval(e_1) = s_1$  and
  $eval(e_2) = s_2$. $e_1 \join e_2$ is again reachable, so since $T$ is
  \Iconfluent{}, $I(e_1 \join e_2)$ which implies $I(s_1 \join s_2)$. Thus, $O$
  is \Iclosed.
\end{elidableproof}

>>>>

\begin{claim}\clmlabel{StateBasedSufficient}
  Given a state-based object $O = (S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$, if $O$ is \Iclosed{}, then $T$ is \Iconfluent.
\end{claim}
\begin{elidableproof}
  Let $e_1, e_2$ be expressions such that $\Irec{e_1}$ and $\Irec{e_2}$. Let
  $s_1 = eval(e_1)$ and $s_2 = eval(e_2)$. Then, $I(s_1)$ and $I(s_2)$, so
  $I(s_1 \join s_2)$ which implies $I(e_1 \join e_2)$. Thus, by
  \clmref{StateBasedTwoIconfluenceDefs}, $T$ is \Iconfluent{}.
\end{elidableproof}

\begin{claim}\clmlabel{OpBasedSufficient}
  Given a operation-based object $O = (S, s_0)$, a set of transactions $T$, and
  an invariant $I$, if $O$ is \Iclosed{}, then $T$ is \Iconfluent.
\end{claim}
\begin{elidableproof}
  Let $t \in T$ and $e_1, e_2$ be expressions such that $\Irec{e_1}$,
  $\Irec{e_2}$, and $I(t(e_1)(e_1))$. Let $s_1 = eval(e_1)$ and $s_2 =
  eval(s_2)$. Then, $I(s_1)$, $I(s_2)$, and $I(t(s_1)(s_1))$, so
  $I(t(s_1)(s_2))$. This implies $I(t(e_1)(e_2))$, so by
  \clmref{OpBasedTwoIconfluenceDefs}, $T$ is \Iconfluent.
\end{elidableproof}

As a special case, \clmref{StateBasedSufficient} tells us that a set $T$ of
transactions is \Iconfluent{} if $I$ is monotonic and $O$ is a semilattice.

\begin{claim}\clmlabel{Monotonicity}
  Let $O = (S, s_0, \join)$ be a state-based object where $(S, \join)$ is a
  semilattice. Let $s \leq s' \defeq s \join s' = s'$. Let $T$ be an arbitrary
  set of transactions. Let $I$ be a \defword{monotonic}: for all $s, s'\in S$,
  if $s \leq s'$ and $I(s)$, then $I(s')$. Also assume $I(s_0)$. $T$ is
  \Iconfluent{}.
\end{claim}
\begin{elidableproof}
  Let $s_1$ and $s_2$ be arbitrary states such that $I(s_1)$ and $I(s_2)$. $s_1
  \leq s_1 \join s_2$, so by monotonicity, $I(s_1 \join s_2)$. Thus, by
  \clmref{StateBasedSufficient}, $T$ is \Iconfluent{}.
\end{elidableproof}

Unfortunately, the converses of \clmref{StateBasedSufficient} and
\clmref{OpBasedSufficient} are not true. Being \Iconfluent{} does not
necessarily imply being \Iclosed{}, as the next two examples show.

\begin{example}\examplelabel{StateBasedNotNecessary}
  Consider the state-based object $O = (S, s_0, \join)$ where $(S, \join)$ is
  $(\ints, \max) \times (\ints, \max)$ and $s_0 = (0, 0)$. Let $T$ consist of
  two transactions $t_{x+1} \defeq \lambda (x, y).\ (x + 1, y)$ and $t_{y-1} =
  \lambda (x, y).\ (x, y - 1)$. Let $I = \setst{(x, y)}{xy \leq 0}$.
  %
  $\setst{(x, y)}{\reachable{(x, y)}} = \nats \times -\nats \subseteq I$, so
  $T$ is \Iconfluent. However, let $s_1 = (-1, 1)$ and $s_2 = (1, -1)$.
  $I(s_1)$ and $I(s_2)$. However, $s_1 \join s_2 = (1, 1)$, and $\lnot I((1,
  1))$. Thus, $O$ is not \Iclosed{}.
\end{example}

@@

Let's take a closer look at \exampleref{StateBasedNotNecessary} (illustrated in
\figref{StateBasedNotNecessary}) to understand why being \Iconfluent{} doesn't
imply being \Iclosed{}. Our counterexample considers points $s_1 = (-1, 1)$ and
$s_2 = (1, -1)$. $s_2$ is reachable, but $s_1$ is not. This is not a
coincidence! If we consider an \Iconfluent{} set of transactions, then any two
reachable points satisfy the invariant, and the join of the two points is again
reachable, so it must satisfy the invariant as well. The only way to join two
invariant satisfying points and end up not satisfying the invariant is if one
or both points are not reachable. Thus, if we know that $I$ is a subset of
reachable points, then being \Iconfluent{} and being \Iclosed{} are equivalent.

\begin{claim}\clmlabel{SufficientAndNecessary}
  If every invariant satisfying state is reachable, then being \Iconfluent{}
  and being \Iclosed{} are equivalent. Being \Iclosed{} is always sufficient,
  but if we can establish that every invariant satisfying point is reachable,
  then being \Iclosed{} is a both necessary and sufficient for being
  \Iconfluent{}.
  \[
    I \subseteq \setst{e}{\reachable{e}} \implies
    (\text{\Iclosed{}} \iff \text{\Iconfluent})
  \]
\end{claim}
\begin{elidableproof}
  Consider a state-based object $O = (S, s_0, \join)$, a set of transactions
  $T$, and an invariant $I$ where $I \subseteq \setst{e}{\reachable{e}}$.
  %
  Being \Iclosed{} always implies being \Iconfluent{} whether or not $I
  \subseteq \setst{e}{\reachable{e}}$. Thus, we only have to prove that if $T$
  is \Iconfluent{}, then $O$ is \Iclosed. If $T$ is \Iconfluent{}, then
  $I(s_0)$. Next, let $s_1$ and $s_2$ be arbitrary states such that $I(s_1)$
  and $I(s_2)$. Both states satisfy the invariant, so both are reachable. That
  is, there exist expressions $e_1$ and $e_2$ where $eval(e_1) = s_1$  and
  $eval(e_2) = s_2$. $e_1 \join e_2$ is again reachable, so since $T$ is
  \Iconfluent{}, $I(e_1 \join e_2)$ which implies $I(s_1 \join s_2)$. Thus, $O$
  is \Iclosed.
\end{elidableproof}
